An Efficient Forecasting Approach\\ to Reduce Boundary Effects in Real-Time Time-Frequency Analysis

### 2021-01-28 Paper/response/response.tex

\item
\textit{page 7, second column. The authors wrote that the choice of the $M$ parameter is especially crucial when the the deterministic component of the signal is no longer stationary, like in the AHM model.
However, when they present an application of the proposed method to a signal satisfying the AHM model, neither they discuss how they select the $M$ value, nor they provide the analysis of the dependency of the performance of the proposed algorithm on the M value. Please include in section IV-C2 the evolution of the performance of the proposed method as a function of M and an explanation on how to tune its value in applications.}\\
\textbf{We have added the new paragraph IV-C2a where we discuss the influence of the parameter $M$ on the forecasting performance of an adaptive harmonic signal. In particular, we mention that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily, while a too large $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.}

@@

\textbf{This results have been added to Table III.}
\item
\textit{All real life signals proposed in the work are nonstationary signals which contain one single nonstationary component plus noise. In the toy example of a  AHM signal, section IV-C2, you presented the case of a signal containing two nonstationary components. Please add the example of a real life signal containing two or more simple nonstationary components plus noise.}\\
\textbf{Section IV-D2 now includes the application of \textsf{BoundEffRed} to a respiratory signal that contains not only the respiratory cycle, but also a
cardiac component, knowm as the cardiogenic artifact. We then show the ability of our algorithm to work on signals containing several nonstationary components. Another application to an atrial blood pressure signal is provided in section III of the Supplementary Materials. This signal also contains both simultaneously a cardiac component and a respiratory component.}

@@

\end{enumerate}

\section*{Reviewer 2}
Thanks for your careful evaluation of the paper and your comments, which we tried to take into account as much as we could. You will find below answers to your remarks, suggestions and questions.
\begin{enumerate}[1)]
\item
\textit{The approach is similar to the shadowing approach suggested by C. Gregobi in the Nonlinear Dynamics community, (see his Physics Review Letters in 1990 Shadowing of physical trajectories in chaotic dynamics) and subsequent papers on shadowing in multiple domains e.g. by Vargas et al in IEEE Transactions on Audio, Speech and Language in 2011.}\\
\textbf{We have mentioned this approach in the introduction of our article. We precise that dynamic mode predictors have been proposed in speech processing by Vargas and McLaughlin, to forecast signals  falling into the so-called source--filter model. This gradient-based technique rely on the shadowing approach proposed in Grebogi et al. Nevertheless,  since the source--filter is not adapted to model physiological signals on which we focus, we do not implement this method in our article.}
\item
\textit{The signal is assumed to be deterministic with additive noise, in my experience oscillatory signals of interest in many applications are a combination of deterministic and stochastic components. How robust is the scheme?}\\
\textbf{A new paragraph has been added to precise the limitations of the theoretical justifications on the performance of \textsf{SigExt}. Indeed, the models of sections III-A and III-C consider the meaningful part of the signal to be a deterministic component. These models are purposely adapted to signals showing local line spectra. The physiological signals we are interested in, such as respiratory or cardiac signals, have this characteristic (see section IV-D). Signals with wider local spectra, such as electroencephalography signals, do not fall into this category, and are more faithfully modeled as random signals. Thus, the theoretical justifications proposed above are no longer applicable to guarantee the forecasting quality of \textsf{SigExt}. A more robust approach to the study of signals following this model is envisaged in our future work.}
\item
\textit{The reference to Takens is slightly confusing could this be explained more clearly. You are not referring to nonlinear dynamic signals so I am not sure it is relevant.} \\
\textbf{Reviewer's comment is judicious. The paragraph comparing the linear dynamical model that we propose with the Takens' embedding wasn't appropriate. For the sake of clarity, this paragraph has been removed.}
\item
\textit{The paper discusses asymptotic behaviour of the forecasting algorithm but the time series lengths in practice are quite short so what impact does that have on performance.} \\
\textbf{In practice, the asymptotic behavior of \textsf{SigExt}, illustrated on a synthetic signal in the paper, can be generalized to the study of real-life signals, as long as the duration of the recorded signal largely exceeds that of the desired extension, which allows the user to set $K \gg L$. In other words, to ensure the relevance of the extension produced by \textsf{SigExt}, the user should be able to use the previously recorded signal for at least three times the duration of the desired extension, approximately. For example, in section IV-D3, the analysis of the PPG signal requires 7-second-long extensions to effectively reduce boundary effects in the TF domain. About twenty seconds of recording are thus necessary to ensure the fidelity of our algorithm.}
\end{enumerate}

### 2021-01-28 Paper/BoundEffRed.tex
polish

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~seconds. A video of these real-time SSTs is available online at \url{https://github.com/AdMeynard/BoundaryEffectsReduction}}
\label{fig:ex.intro}
\end{figure}

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
% 
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique rely on the shadowing approach proposed in~\cite{Grebogi90shadowing}. 
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

Thus, assuming that the quality of the extension step is sufficient, the analysis results obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the nonextended signal. We claim and prove that forecasting oscillatory signals based on the simple dynamical model combined with the simple least square approach is sufficient for reducing the boundary effects for the TF analysis, or other kernel-based analysis. To our knowledge, such theoretical analysis does not exist in the literature. See the bottom right of Fig.~\ref{fig:ex.intro} for a snapshot of the result. The main benefit of this simple approach is a numerically efficient solution with a theoretical guarantee for real-time analysis purposes.

>>>>

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~seconds. Videos of these real-time SSTs are available online at \url{https://github.com/AdMeynard/BoundaryEffectsReduction}}
\label{fig:ex.intro}
\end{figure}

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
%
{\color{red}
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique rely on the shadowing approach proposed in~\cite{Grebogi90shadowing}.
}
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

Thus, assuming that the quality of the extension step is sufficient, the analysis results obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the nonextended signal. We claim and prove that forecasting oscillatory signals based on the simple dynamical model combined with the simple least square approach is sufficient for reducing the boundary effects for the TF analysis, or other kernel-based analysis. To our knowledge, such theoretical analysis does not exist in the literature. See the bottom of Fig.~\ref{fig:ex.intro}---in particular, the regions indicated by the dashed boxes---for a snapshot of the result. The main benefit of this simple approach is a numerically efficient solution with a theoretical guarantee for real-time analysis purposes.

### 2021-01-28 Paper/Supplementary material/SupMat.tex
polish

\setcounter{equation}{25}
\setcounter{figure}{8}
\setcounter{table}{3}

@@

\section{Application to a Multicomponent Cardiac Signal}
We consider a cardiac signal, namely an arterial blood pressure recording, sampled at $1000$~Hz. In addition to the blood pressure measurement, this signal contains a slow varying component, which is the respiratory component. Fig.~\ref{fig:abp} displays the ordinary and boundary-free SSTs of a 150-second-long extract of this signal. This TF analysis brings out both components---the cardiac component, the most energetic, is located around the fundamental frequency $1$~Hz and its multiples, while the respiratory component, less visible, is located around $0.2$~Hz. The performance index of this representation takes the value $D=0.850$. Note that, here, boundary effects are only reduced on the right-side boundary of the TF domain. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $15\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{ABPsst.eps}
\caption{Ordinary SST (left) and boundary-free SST (right) of an arterial blood pressure recording. The red dashed rectangle highlights the
area where boundary effects may appear.}
\label{fig:abp}
\end{figure}

>>>>

\setcounter{equation}{25}
\setcounter{figure}{9}
\setcounter{table}{3}

@@

{\color{red}
\section{Application to a Multicomponent Cardiac Signal}
We consider a cardiac signal, namely an arterial blood pressure (ABP) recording, sampled at $1000$~Hz. In addition to the blood pressure measurement, this signal contains a slow varying component, which is the respiratory component. The top Fig.~\ref{fig:abp} displays an excerpt of the signal along with a 7-second-long extension obtained by the {\sf SigExt} forecasting. The forecasting breaks the waveform of the oscillations because of its inability to forecast the high-frequency harmonics contained in the signal. Nevertheless, as long as the low-frequency harmonics of the components contained in the signal are preserved, the forecasting is sufficient to reduce boundary effects in the low-frequency TF domain.

The ordinary and boundary-free SSTs of this signal are displayed in the bottom of Fig.~\ref{fig:abp}. These TF analyses brings out both components---the cardiac component, the most energetic, indicated by the blue arrows, is located around the fundamental frequency $1$~Hz and its multiples, while the respiratory component, less visible, indicated by the green arrows, is located around $0.2$~Hz. Clearly, near the boundaries, {\sf BoundEffRed} helps improve the quality of the TF representation. Besides, the performance index of this representation takes the value $D=0.679$. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $32\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{ABP.eps}
\includegraphics[width=\textwidth]{ABPsst.eps}
\caption{Atrial blood pressure signal (top), and the associated
{\sf SigExt} extension (blue). Ordinary SST (bottom left) and boundary-free SST (bottom right) of the arterial blood pressure signal. The window length for the SST is 18~seconds. The green arrows indicate the instantaneous frequency of the respiratory component. The blue arrows, on the other hand, indicate the instantaneous frequency of the blood pressure component. The red dashed boxes highlight the areas where boundary effects may appear.}
\label{fig:abp}
\end{figure} 
}

### 2021-01-28 Paper/algorithm.tex
polish

\paragraph{Dynamical Model} 
Establishing a dynamical model consists of determining the relation linking $\bY$ to $\bX$, that is finding a function $f$ so that
\begin{equation}
\bY = f(\bX)\ .
\label{eq:generic.model}
\end{equation}
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} or other more complicated models \cite{Roberts13Gaussian,DeLivera11forecasting,west2006bayesian,vlachas2018data}, allow this by setting additional constraints on the behavior of $f$. We will see, in section~\ref{se:theoretical}, that considering such a complex dynamic model is not necessary for the study of the oscillatory signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. In other words, we adopt a classical strategy in the study of dynamical systems, the {\em linearization of a nonlinear system}, when the system is sufficiently regular. Notice that this linearized dynamical model can be written equivalently according to the subsignals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}
%
%{\color{blue}
%From the Takens' embedding perspective~\cite{Takens81detecting}, $\bX$ can be viewed as a lag map. Under the manifold assumption of the intrinsic phase space that hosts the dynamics, $f$ describes the dynamics supported on the manifold space, and $\bA$ approximates the dynamics on the manifold.
%}

\paragraph{Forecasting}
The forecasting method consists in estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$. 
%
We then obtain the forecasting of $\bx_{k+\ell}$ by recursively applying the linear relation~\eqref{eq:dyn.model}:
\begin{equation*}
\tilde\bx_{K+\ell} = \underbrace{\tilde\bA\tilde\bA\cdots\tilde\bA}_{\scriptstyle\ell\mbox{ \footnotesize {times}}}\bx_{K} = \tilde\bA^\ell\bx_{K}\ .
\end{equation*}  
Let $\be_M$ denote the unit vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$. Consequently, given definition~\eqref{eq:xk}, the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ is provided by
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \be_M^T\tilde\bx_{K+\ell} = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$, \ie~$\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell$.

>>>>

\paragraph{Dynamical Model} 
Establishing a dynamical model means determining the relation linking $\bY$ to $\bX$; that is, finding a function $f$ so that
\begin{equation}
\bY = f(\bX)\ .
\label{eq:generic.model}
\end{equation}
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} or other more complicated models \cite{Roberts13Gaussian,DeLivera11forecasting,west2006bayesian,vlachas2018data}, allow this by setting additional constraints on the behavior of $f$. We will see, in section~\ref{se:theoretical}, that considering such a complex dynamic model is not necessary for the study of the oscillatory signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. In other words, we adopt a classical strategy in the study of dynamical systems, the {\em linearization of a nonlinear system}, when the system is sufficiently regular. Notice that this linearized dynamical model can be written equivalently according to the subsignals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}
%
%{\color{blue}
%From the Takens' embedding perspective~\cite{Takens81detecting}, $\bX$ can be viewed as a lag map. Under the manifold assumption of the intrinsic phase space that hosts the dynamics, $f$ describes the dynamics supported on the manifold space, and $\bA$ approximates the dynamics on the manifold.
%}

{\color{red}
\paragraph{Forecasting}
Determining the forecasting amounts to estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$. 
%
We then obtain the forecasting of $\bx_{k+\ell}$ by recursively applying the linear relation~\eqref{eq:dyn.model}:
\begin{equation*}
\tilde\bx_{K+\ell} = \underbrace{\tilde\bA\tilde\bA\cdots\tilde\bA}_{\scriptstyle\ell\mbox{ \footnotesize {times}}}\bx_{K} = \tilde\bA^\ell\bx_{K}\ .
\end{equation*}  
Let $\be_M$ denote the unit vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$. Consequently, given definition~\eqref{eq:xk}, the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ is provided by
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \be_M^T\tilde\bx_{K+\ell} = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$, \ie~$\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell$.
}

### 2021-01-28 Paper/results.tex
polish

As anticipated by Theorem~\ref{th:error}, the variance of the estimator decreases linearly with $\frac1K$ as long as the dataset size is large enough. Indeed, in the log-log plot displayed in Fig.~\ref{fig:res.size.sine}, as soon as $K>700$, each of the three curves representing the forecasting variance as a function of $K$ tends towards an asymptote of slope $-1$. 
%
In practice, the asymptotic behavior of {\sf SigExt}, illustrated here on a synthetic signal, can be generalized to the study of real-life signals, as long as the duration of the recorded signal largely exceeds that of the desired extension, which allows the user to set $K\gg L$.

@@

\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to three different values of the subsignals length $M$. Note that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a too large $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.

@@

Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also the cardiogenic artifact which reflects the hemodynamic component~\cite{Smith94recognition}. The top of Fig.~\ref{fig:resp.2} displays the ordinary SST and the boundary-free SST of a 180-second-long extract of this signal, sample at $125$~Hz. This TF analysis brings out both components---the respiratory component, the most energetic, is located around the fundamental frequency $0.25$~Hz and its multiples, while the cardiac component, less visible, is located around $1.7$~Hz. Note that, here, boundary effects are only reduced on the right-side boundary of the TF domain. The middle of Fig.~\ref{fig:resp.2} show a zoom near the right-boundary of the ordinary SST, while the bottom of Fig.~\ref{fig:resp.2} respectively show the respective zoom on the boundary-free SST. In this area, the boundary-free SST makes it possible to disentangle the components contained in the signal. Moreover, the performance index of this representation takes the value $D=0.314$. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $69\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\includegraphics[width=.48\textwidth]{Resp2compZOOM.eps}
\caption{Ordinary SST (top-left) and boundary-free SST (top-right) of a two-component respiratory signal. Middle and bottom: zoom on the areas delimited by the red dashed rectangles; these images are rotated 90 degrees clockwise.}
\label{fig:resp.2}
\end{figure}

>>>>

As anticipated by Theorem~\ref{th:error}, the variance of the estimator decreases linearly with $\frac1K$ as long as the dataset size is large enough. Indeed, in the log-log plot displayed in Fig.~\ref{fig:res.size.sine}, as soon as $K>700$, each of the three curves representing the forecasting variance as a function of $K$ tends towards an asymptote of slope $-1$. 
%
{\color{red}
In practice, the asymptotic behavior of {\sf SigExt}, illustrated here on a synthetic signal, can be generalized to the study of real-life signals, as long as the duration of the recorded signal largely exceeds that of the desired extension, which allows the user to set $K\gg L$.
}

@@

{\color{red}
\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to three different values of the subsignals length $M$. Note that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a too large $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.
}

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
%
{\color{red}
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.
}

@@

{\color{red}
\subsubsection{Two-Component Respiratory Signal}
Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also a cardiac component known as the cardiogenic artifact~\cite{Smith94recognition}. The top of Fig.~\ref{fig:resp.2} shows an excerpt of the measured signal, as well as a 10-second-long {\sf SigExt} extension. Below, the simultaneously recorded ECG signal emphasizes the coincidences between the cardiogenic artifact of the respiratory signal and the ECG.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2.eps}
\caption{Two-component respiratory signal (top), and the associated {\sf SigExt} extension (blue); the simultaneously recorded ECG signal is below.}
\label{fig:resp.2}
\end{figure}

The top of Fig.~\ref{fig:resp.2.sst} displays the ordinary SST and the boundary-free SST of this signal, sampled at $125$~Hz. This TF analysis brings out both components---the respiratory component, the most energetic, is located around the fundamental frequency $0.25$~Hz and its multiples, while the cardiac component, less visible, is located around $1.7$~Hz. Note that, here, boundary effects are only reduced on the right-side boundary of the TF domain. The middle of Fig.~\ref{fig:resp.2.sst} show a zoom near the right-boundary of the ordinary SST, while the bottom of Fig.~\ref{fig:resp.2.sst} respectively show the respective zoom on the boundary-free SST. In this area, the boundary-free SST makes it possible to disentangle the components contained in the signal. Moreover, the performance index of this representation takes the value $D=0.205$. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $80\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\includegraphics[width=.48\textwidth]{Resp2compZOOM.eps}
\caption{Ordinary SST (top-left) and boundary-free SST (top-right) of a two-component respiratory signal. Middle and bottom: zoom on the areas delimited by the red dashed rectangles; these images are rotated 90 degrees clockwise. The window length for the SSTs is 35~seconds.}
\label{fig:resp.2.sst}
\end{figure}
}
